The Chemistry Development Kit (CDK) \cdkversion{}: atom typing, depiction, molecular formulas, and substructure searching

### 2017-05-01 article.tex
Update numbers and wording in benchmark section.

The improvement on this task is most noticeable for SMILES input, previously
it would take more than 8 minutes to read ChEMBL 22.1 but this is reduced
to only 12 seconds. On top of this improvement SMILES input is now validated 
and assigned a Kekul$\mathrm{\acute{e}}$ structure. This identifies 9 invalid
entries in ChEBI and another 9 in ChEMBL. Most of these rejected SMILES
are due to the wrong encoding of Cis/Trans double bond stereochemistry at ring 
closures. The ChEBI 149 SMILES input has 2107 empty records that v1.4.19 skip,
\cdkversion{} simply reads these as empty molecules. Input from SDF also 
improved from $\sim$3 minutes to $\sim$1 minute for ChEMBL. The SDF input in
\cdkversion{} now includes perception of stereochemistry and reading CTfile 
Sgroups. There are 9 entries from ChEBI's SDF that are rejected because they 
contain CTfile query features (e.g. any bond order).

@@

\textbf{-mark}, The first subtask measures the performance in marking 
ring membership and reporting the number of ring bonds in each record.
This requires a linear algorithm based on a depth first search. 
The original code used a weighted spanning tree to compute the membership in 
linearithmic time. The run times are similar for these datasets (Table
7), larger differences are seen for more complex cage molecules such fullerenes
\cite{May2014}.

@@

\textbf{-all}, The third subtask counts the number of all rings up to or equal
to size 12. This includes rings that encompass other smaller rings, for 
example, 1H-indole has rings of size 5, 6, and 9. In general this problem is 
exponential and so an adjustable threshold or timeout is used to avoid problematic 
molecules. CDK v1.4.19 used a timeout based threshold (default 5 seconds) 
whilst \cdkversion{} uses a 
counter based on properties of algorithm~\cite{May2014}. In \cdkversion{} there
were 15/17\footnote{2 records from SDF use query bond features and are skipped
when read} records skipped from ChEBI that have complex cage-like ring systems 
(e.g. CHEBI:33611), no records in ChEMBL reached the threshold. By comparison in
v1.4.19 there were 14/16 records skipped from ChEBI and 88/90 in ChEMBL due to
reaching the time out. The number fluctuates here as a timeout based threshold 
depends on external factors such as CPU load.

The speed-up in \cdkversion{} is slightly better than the SSSR task. ChEMBL 
previously spent 4-5 minutes and now takes only $\sim$14 seconds (Table 6).
In \cdkversion{} finding all rings ($\leq$ 12 bonds) runs faster than the 
non-unique SSSR computation.

@@

This task measures the generation of a Unique SMILES string. These can be
used to compare dataset intersection and exact lookup. From SMILES input
\cdkversion{} the total elapsed time is $\sim$20 times faster for both ChEBI
and ChEMBL. For ChEMBL it now takes just under 43 seconds to read, reorder, 
and write the SMILES compared to more than 14 minutes previously.

@@

\textbf{-ofmt smi}: SMILES is a very compact means of storing connection tables, v1.4.19 could 
only write canonical SMILES, \cdkversion{} allows different SMILES flavours 
to be generated including a non-canonical variant. This task outputs
CXSMILES that includes additional fields such as repeat groups (used by some 
ChEBI entries). As
expected the v1.4.19 execution time is the same as for the Canonical SMILES 
task but \cdkversion{} can generate the non-canonical SMILES faster taking less
than 30 seconds for ChEMBL.

@@

\textbf{-gen2d -ofmt sdf}: When writing SDF the only portable way to store stereochemistry is with the 
inclusion of coordinates, this is specified with the \texttt{-gen2d} option. 
The overhaul in layout generation discussed early provides better layouts but 
also included performance tweaks, in CDK v1.4.19 generating coordinates and 
writing an SDF for ChEMBL took almost 3.5 hours but now only takes just over
18 minutes.

@@

\textbf{-type path}: Path based fingerprints encode paths of length 0 to 7. Path
based fingerprints can be used for both substructure and similarity screening. 
The algorithm was tweaked for \cdkversion{} to hash the paths without 
pre-computing all paths upfront and without needing to generate character 
strings before hashing. The time to encode ChEMBL previously took 42-47 minutes 
now only takes 6-7 minutes.

\textbf{-type maccs}: The CDK MACCS fingerprint uses 166 keys to encode
features of a structure and can be used for similarity searching. This encoding
uses different aspects of the library including ring perception and the new
aromaticity perception but the speed is primarily dependent on SMARTS matching 
performance. Generating the fingerprint previously took almost 1 day for ChEMBL 
and 2 hours for ChEBI. This has been reduced to less than 
15 minutes for ChEMBL and less than 30 seconds for ChEBI.

>>>>

The improvement on this task is most noticeable for SMILES input, previously
it would take more than 8 minutes to read ChEMBL 22.1 but this is reduced
to less than 11 seconds. On top of this improvement SMILES input is now validated 
and assigned a Kekul$\mathrm{\acute{e}}$ structure. This identifies 9 invalid
entries in ChEBI and another 9 in ChEMBL. Most of these rejected SMILES
are due to the wrong encoding of Cis/Trans double bond stereochemistry at ring 
closures. The ChEBI 149 SMILES input has 2107 empty records that v1.4.19 skip,
\cdkversion{} simply reads these as empty molecules. Input from SDF also 
improved from $\sim$3 minutes to $\sim$1 minute for ChEMBL. The SDF input in
\cdkversion{} now includes perception of stereochemistry and reading CTfile 
Sgroups. There are 9 entries from ChEBI's SDF that are rejected because they 
contain CTfile query features (e.g. any bond order).

@@

\textbf{-mark}, The first subtask measures the performance in marking 
ring membership and reporting the number of ring bonds in each record.
This requires a linear algorithm based on a depth first search. 
The original code used a weighted spanning tree to compute the membership in 
linearithmic time. The run times are similar for these datasets (Table
7), larger differences are only seen for more complex cage molecules such 
fullerenes~\cite{May2014}.

@@

\textbf{-all}, The third subtask counts the number of all rings up to or equal
to size 12. This includes rings that encompass other smaller rings, for 
example, 1H-indole has rings of size 5, 6, and 9. In general this problem is 
exponential and so an adjustable threshold or timeout is used to avoid problematic 
molecules. CDK v1.4.19 used a timeout based threshold (default 5 seconds) 
whilst \cdkversion{} uses a 
counter based on properties of algorithm~\cite{May2014}. In \cdkversion{} there
were 15/17\footnote{2 records from SDF use query bond features and are skipped
when read} records skipped from ChEBI that have complex cage-like ring systems 
(e.g. CHEBI:33611), no records in ChEMBL reached the threshold. By comparison in
v1.4.19 there were 14/16 records skipped from ChEBI and 88/90 in ChEMBL due to
reaching the time out.

The speed-up in \cdkversion{} is slightly better than the SSSR task. ChEMBL 
previously spent 4-5 minutes and now takes only $\sim$12-14 seconds (Table 6).
In \cdkversion{} finding all rings ($\leq$ 12 bonds) runs faster than the 
non-unique SSSR computation.

@@

This task measures the generation of a Unique SMILES string. These can be
used to compare dataset intersection and exact lookup. From SMILES input
\cdkversion{} the total elapsed time is $\sim$20 times faster for both ChEBI
and ChEMBL. For ChEMBL it now takes just under 41 seconds to read, reorder, 
and write the SMILES compared to more than 14 minutes previously.

@@

\textbf{-ofmt smi}: SMILES is a very compact means of storing connection tables, v1.4.19 could 
only write canonical SMILES, \cdkversion{} allows different SMILES flavours 
to be generated including a non-canonical variant. This task outputs
CXSMILES that includes additional fields such as repeat groups (used by some 
ChEBI entries). As
expected the v1.4.19 execution time is the same as for the Canonical SMILES 
task but \cdkversion{} can generate the non-canonical SMILES faster taking less
than 30 seconds for SMILES from ChEMBL.

@@

\textbf{-gen2d -ofmt sdf}: When writing SDF the only portable way to store 
stereochemistry is with the inclusion of coordinates, this is specified with 
the \texttt{-gen2d} option. The overhaul in layout generation discussed early 
provides better layouts but also included performance tweaks, in CDK v1.4.19 
generating coordinates and writing an SDF for ChEMBL took almost 3.5 hours but
now only takes $\sim$18 minutes.

@@

\textbf{-type path}: Path based fingerprints encode paths of length 0 to 7. Path
based fingerprints can be used for both substructure and similarity screening. 
The algorithm was tweaked for \cdkversion{} to hash the paths without 
pre-computing all paths upfront and without needing to generate character 
strings before hashing. The time to encode ChEMBL previously took 42-47 minutes 
now only takes 6-8 minutes.

\textbf{-type maccs}: The CDK MACCS fingerprint uses 166 keys to encode
features of a structure and can be used for similarity searching. This encoding
uses different aspects of the library including ring perception and the new
aromaticity perception but the speed is primarily dependent on SMARTS matching 
performance. Generating the fingerprint previously took $\sim$1 day for ChEMBL 
and $\sim$1.75 hours for ChEBI. This has been reduced to less than 
13.5 minutes for ChEMBL and $\sim$20 seconds for ChEBI.
